---
title: "Ernst Cemetery Example"
author: "MDH"
date: "September 25, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ernst-Witte Site Burials Analysis

This example runs through the exploratory data analysis (EDA) of the Ernst-Witte burials data.  Following this is the creation of a presentable plot of percent pf grave goods over individual age by time period.  Dplyr data munging is used to transform from the original dat a to a format the is useful for this plot.

This data set includes sex, age, burial group, location, and burial orientation and direction facing from the Ernest Witte
site, a Late Archaic cemetery in Texas (Hall 1981).

#### Note about ggplot2 version
this relies on the development version of ggplot2 for the subtitle and caption text.  To install this version, first install the `devtools` package `install.packages("devtools")`, then install ggplot from github `devtools::install_github("hadley/ggplot2")`.  If you do not do this and have the CRAN version, everything should work except the subtitle and caption, and you will get a warning.


### Load packages
```{r libraries, comment=''}
library("archdata")
library("dplyr")
library("tidyr")
library("ggplot2") # dev version
library("scales")
```

### assign to `ew` object
```{r load_data, comment=''}
data(EWBurials)
ew <- EWBurials
```

### Explore data structure
Common use of `str()`, `summary()`, and `head()`
```{r eds, comment=''}
str(ew)
summary(ew)
head(ew)
```


### Data processing
The purpose of this data processing is to go from the wide data format to a long dat format that answers the question: what is the percent of grave goods by male/female for each burial individual age and period of burial.

The process below follows:
  * `group_by` to group the data by Age, Sex, and Group (period)
  * `mutate` the data to change the values of `Goods` to presence/absence
  * `summarise` the data to `sum` the presence of Goods
  * count the sum of Goods with `n()`
  * and calculate the percent of grave goods
  * the data are `ungroup`'ed 
  * finally, `complete` creates all potential combinations of Age, Sex and Group and fills in those without a percent with a `0`.
  
The final step using `complete` makes piloting more consistent.
```{r dplyr, comment=''}
ew2 <- group_by(ew, Age, Sex, Group) %>%
  mutate(Goods = ifelse(Goods == "Present", 1, 0)) %>%
  summarise(sum_goods = sum(Goods),
            cnt = n(),
            pcnt = sum_goods/cnt) %>%
  ungroup() %>%
  tidyr::complete(Age, Sex, Group, fill = list(pcnt = 0))
```

#### Preliminary plot
Plot the data from the dplyr sequence for a sanity check
```{r initial_plot, comment=''}
ggplot(ew2, aes(x = Age, y = pcnt, group = Sex, fill = Sex)) + 
  geom_bar(stat = "identity", position = "dodge")
```

### Build ggplot
Once the data is worked into the format desired to answer the question we are interested in, we can build a ggplot to display the results.

The following plot uses the long dat format created above, uses the `facet_grid()` function to split the plots into two broad time periods, `scale_fill_manual()` to set custom fill colors, a bunch of calls to `theme()` to adjust plot elements, and finally within `theme()` uses `legend.position()` to move the legend to a spot within the plot.

In less than 100 lines of code, we took data from its original format, transformed it to address a specific question, and then made a publication ready plot.

```{r final_plot, comment=''}
group_names <- c(`1` = " 2000 - 1200 BCE", `2` = " CE 200 - 500")

ggplot(ew2, aes(x = Age, y = pcnt, group = Sex, fill = Sex)) + 
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 0, color = "gray70") +
  theme_bw() +
  scale_fill_manual(values = c("darkgoldenrod2", "slategrey")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title="Percent of Grave Goods by Age, Sex, and Temporal Group",
       subtitle="Ernest Witte site, Austin County, Texas",
       caption="Data: Hall,G.D. (1981)",
       y = "Grave Good Presence") +
  facet_grid(Group~., labeller = as_labeller(group_names)) +
  theme(
    strip.background = element_rect(colour = "white", fill = "white"),
    strip.text.y = element_text(colour = "black", size = 10, face = "bold", 
                                family = "Trebuchet MS"),
    panel.border = element_rect(color = "gray90"),
    axis.text.x = element_text(angle = 0, size = 10, family = "Trebuchet MS"),
    axis.text.y = element_text(size = 9, family = "Trebuchet MS"),
    axis.title.y = element_text(size = 11, family = "Trebuchet MS"),
    axis.title.x = element_blank(),
    plot.caption = element_text(size = 10, hjust = 0, margin=margin(t=10), 
                                family = "Trebuchet MS"),
    plot.title=element_text(family="TrebuchetMS-Bold"),
    plot.subtitle=element_text(family="TrebuchetMS-Italic"),
    panel.grid.major.x = element_blank(), 
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = c(0.1, 0.85),
    panel.spacing = unit(1, "lines")
  )
```


## Data set information

#### Details
The Ernest Witte site in Austin County, Texas contains four burial groups from different time periods.
Group 1 includes 60 interments and that occurred between about 2000 and 1200 BCE. Group
2 is the largest with 148 interments. The burials in this group were interred between about CE 200
and 500. Groups 3 and 4 include only 10 and 13 interments and date to CE 500 to 1500, but are
not included in this data set which was taken from Appendix II (Hall 1981). Two of the variables,
direction and looking, are circular data and require package circular. Hall (2010) provides a
summary of the site and its significance.


#### Source
Hall, G. D. 1981. Allen’s Creek: A Study in the Cultural Prehistory of the Lower Brazos River
Valley. The University of Texas at Austin. Texas Archaeological Survey. Texas. Research Report No.
61.


#### References
Hall, G. D. 2010. Ernest Witte site. Handbook of Texas Online http://www.tshaonline.org/
handbook/online/articles/bbe05. Texas State Historical Association


