---
title: "Mapping Spatial Data in R"
author: "MDH"
date: "September 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mapping Spatial Data

This markdown document shows some of the methods for importing and creating spatial dat in R and mapping these data in both static and interactive forms.  There is a ton of great information in R mapping/GIS on the web.  The authoritative source for keeping track of what exists is the [CRAN Task View: Analysis of Spatial Data](https://cran.r-project.org/web/views/Spatial.html).  Googling more specific packages will certainly get you some valuable information.  Essentially, R is a fully functional GIS by any definition of the phrase.  However, it is not a replacement for everything you may do in ArcGIS or QGIS, but it is a replacement for many tasks and an extension of these other software.  Mapping in R really makes gains in spatial modeling, handling large data sets, repeatable tasks, and working spatial data into a larger analysis

The example here uses the `Michelsberg` data set from the `archdata` package.  These data include the location and attributes of vessels gathered from 108 Michelsberg period features.  The code below walks through how to make these data into spatial objects, how to import a shapefile, create a static map, conduct spatial analysis on a raster, and finally map these layers in an interactive map.

#### Data details:
A sites by types table of abundance data on vessel types in archaeological features of the Younger Neolithic Michelsberg Culture from Belgium, France and Germany by Birgit Höhn (2002).

### Load libraries
```{r libraries, message=FALSE, warning=FALSE, comment=''}
library("archdata") # where the data comes from
library("sp") # for spatial object
library("leaflet") # for interactive mapping
library("rgdal") # for geographic transformations
library("rgeos") # for geographic operations
library("spatstat")  # Needed for the dirichlet tesselation function
library("rworldmap") # library rworldmap provides different types of global maps, e.g:
library("ggmap") # for a static map example
library("gstat") # for IDW model
library("leaflet") # for interactive mapping
library("mapview") # for interactive mapping test with mapview(breweries91)
library("raster") # for raster operations

```


### Invoke data and note structure
```{r load data, comment=''}
### bring in data from archdata package
data(Michelsberg)
# give a shorter name, less typing...
mb <- Michelsberg
# take a look at structure of data
str(mb)
# take a look at type and range of variables
summary(mb)
```

## Making data spatial

### create objects for coordinate referrence systems
```{r CRS, comment=''}
## Proj4string from http://spatialreference.org/
# UTM 1983 zone 32 North - Michelsberg data
UTM32n <- CRS("+proj=utm +zone=32 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")
# World Geographic System 1984 (lat/long) - mapping
WGS84 <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84") 
```

### Create a spatialpointsdataframe object from our data
```{r sp_object, comment=''}
## pass long, lat, and data to function.  
## proj4string = the proper CRS as described in `archdata` package reference
mb_spat <-  SpatialPointsDataFrame(coords = mb[,c("x_utm32n", "y_utm32n")], 
                                   data = mb, 
                                   proj4string = UTM32n)
```


### Static mapping with base `plot()`
First make a static map with the base `plot()` function
```{r base_plot, comment=''}
## most basic plot (points)
plot(mb_spat)
```

### Putting it in geographic context
Bring in country boundary data from the `rworldmap` package, reproject the `mb` data to `WGS84` and use base `plot()` again to plot vessel locations, country boundaries, and labels.
```{r geo_plot, comment=''}
# get country boundary data
data(countriesLow)
# transform Michelsberg site data from UTM83n to WGS84
mb_spat_WGS84 <- spTransform(mb_spat, WGS84)
# plot the data and boundaries, then lable
plot(mb_spat_WGS84, pch = 20, col = "red")
# 'add' tell plot to add to the previous plot
plot(countriesLow, add = TRUE)
labelCountries()

```


### ggmap example: Static map with nice basemap
The next approach is to use the `ggmap` package to create nice base maps and use the `ggplot` framework for mapping.

```{r ggmap1, comment=''}
# grab WGS84 coordinates
WGS84_coords <- coordinates(mb_spat_WGS84)
# append WGS84 coords onto original data
mb$x_WGS84 <- WGS84_coords[,1]
mb$y_WGS84 <- WGS84_coords[,2]
# get a base map centered on Frankfurt, Germany at a certain zoom
base_map <- get_map(location = "Frankfurt, Germany", zoom = 6, color = "bw")
# use ggplot code to build out layers of map:
# basemap, add points, change color scale, set plot theme
ggmap(base_map, extent = "normal") +
  geom_point(aes(x = x_WGS84, y = y_WGS84, color = t2a), data = mb, alpha = .8, size = 4) +
  scale_color_distiller(palette = "PuOr", direction = 1) +
  theme_bw(16)
```

## Loading and adding a shapefile to the plot
Loading a shapefile is pretty easy with the `readOGR()` function of the `rgdal` package.  If the shapefile has a crs assigned to it, `readOGR()` will detect it and assign it to the spatial object in R; very handy.  For this demo, we load a shapefile for the border of Germany [in the `/data/` folder on GitHub].  Next we use `gSimplify()` to reduce the number of edges to speed up the rendering. Finally, base `plot()` is used to take a look.

```{r shapefil, comment=''}
wd <- "/Users/mattharris/Documents/R_Local/SAA_R_Demo_2016/SAA_R_Demo_2016"
germany <- readOGR(paste0(wd, "/data/DEU_adm0.shp"), layer = "DEU_adm0")
germany <- gSimplify(germany, tol=0.001, topologyPreserve=TRUE)
plot(germany)
```


### Fortifying for ggplot
To use the imported shapefile in ggmap, we need the `fortify()` function of the `ggplot2` package.  The `fortify()` function converts the spatial data into a dataframe suitable for plotting in `ggmap` or `ggplot`.  As with a typical GIS, the plot is build by layers, first the `geom_polygon()` of the boundary, then the `geom_point()` of the vessel locations, and finally some color and display settings. 

#### Note:
the `ggplot2` function of `coord_map()` is used here to "zoom" the map into the extents of the data.  There are other functions such as `ylim()` and `scale_y_continuous()` that control plot extents, but you will need `coord_map()` to do so with map data.  Otherwise, data is cut off or shifted out of alignment.

```{r fortify_shp, comment=''}
germany_fort <- ggplot2::fortify(germany)
# basemap with points and SHP
ggmap(base_map, extent = "normal") +
  geom_polygon(data = germany, aes(x=long, y=lat, group=group), 
               fill = "transparent", color = "gray10", size = 1) +
  geom_point(aes(x = x_WGS84, y = y_WGS84, color = t2a), 
             data = mb, alpha = .8, size = 4) +
  scale_color_distiller(palette = "PuOr", direction = 1) +
  coord_map(ylim = c(min(mb$y_WGS84), max(mb$y_WGS84)),
              xlim = c(min(mb$x_WGS84), max(mb$x_WGS84))) +
  theme_bw(16)
```


## Spatial Analysis

The first foray into spatial analysis happens directly in out calls to `ggmap()`.  The `ggplot2` framework is so extensive, it even has a number of statistical functions that can manipulate the data directly in the plotting function.  In this case, `stat_bin2d()` is used to aggregate the number of vessels into aerial units.

#### Binned density
```{r analysis, comment=''}
## use ggplot functions to creat site density grid
ggmap(base_map, extent = "normal") +
  stat_bin2d(aes(x = x_WGS84, y = y_WGS84, color = t2a),
  size = 1, bins = 20, alpha = 0.9, data = mb)
```

#### Smooth density
Using the `stat_density2d()` function
```{r smooth_density, comment=''}
ggmap(base_map, extent = "normal") +
  stat_density2d(aes(x = x_WGS84, y = y_WGS84, fill = ..level.., alpha = ..level..),
  size = 1, bins = 20, data = mb,
  geom = "polygon")
```

#### Faceted density
Finally, we create a multi-panel, or faceted, map that shows the data distribution of data filtered on the number of occurrences of vessel type `"t2a"`
#### Note: this can take a while to render!
```{r facetted_density, comment=''}
mb$group <- ifelse(mb$t2a <= 3, "t2a <= 3", "t2a > 3")
ggmap(base_map, extent = "normal") +
  stat_density2d(aes(x = x_WGS84, y = y_WGS84, fill = ..level.., alpha = ..level..),
                 size = 1, bins = 20, data = mb,
                 geom = "polygon") +
  scale_fill_distiller(palette = "PuOr", direction = 1) +
  theme_bw() +
  facet_wrap(~ group)

```

### Analysis with `gstat`
Next we perform a bit more in depth spatial analysis outside of the `ggmap()` call and then layer it back in.

This code uses the `gstat` package to perform an Inverse Distance Weighting (IDW) interpolation on our data (sort of... a new vessel type is created and given a fake distribution that looks better for demonstration purposes).

#### Build a grid to the geographic limits of the data
```{r gstat, comment=''}
# get the min/max range for lat/long to make an empty grid 
x.range <- as.numeric(c(min(mb$x_WGS84), max(mb$x_WGS84)))  # min/max longitude of the interpolation area
y.range <- as.numeric(c(min(mb$y_WGS84), max(mb$y_WGS84)))  # min/max latitude of the interpolation area  
# from the range, exapnd the coordinates to make a regular grid
grd <- expand.grid(x = seq(from = x.range[1], to = x.range[2], by = 0.075), 
                   y = seq(from = y.range[1], to = y.range[2], by = 0.075))  # expand 
```

#### make the grid into spatial coords, grid, assign WGS84 CRS
```{r spatial_grid, comment=''}
coordinates(grd) <- ~x + y
sp::gridded(grd) <- TRUE
proj4string(grd) <- WGS84
# quick plot to see what we have
plot(grd, cex = 1.5, col = "grey")
points(mb_spat_WGS84, pch = 1, col = "red", cex = 1)
```


#### simulate a new vesse type with a nicely defined large scale geographic distribution trend; call it `new_type`
```{r sim_data, comment=''}
### Because most attributes in example data don't have a great distribution
### I am simulating a made up attribute so tha mapping looks nicer
## simulate 'new_type' with spatial trend
# finds mean of site locations,
# assigns a value that increases from lower-left to upper-right
x_mean <- mean(mb$x_WGS84)
y_mean <- mean(mb$y_WGS84)
x_dist <- mb$x_WGS84 - x_mean
y_dist <- mb$y_WGS84 - y_mean
mb$xy_dist <- x_dist + y_dist
mb$pre_sort_id <- seq(1,nrow(mb))
mb <- mb[order(mb$xy_dist),]
mb$new_type <- seq(1,nrow(mb))*2
mb <- mb[order(mb$pre_sort_id),]
mb_spat_WGS84$new_type <- mb$new_type
```

### Perform IDW interpolation
The `gstat::IDW()` function performs the interpolation and then the results are converted into a data.frame using `as.data.frame()`.
```{r IDW, comment=''}
### IDW interpolation of 'new_type' using gstat package
idw <- gstat::idw(formula = new_type ~ 1, locations = mb_spat_WGS84, newdata = grd)  # apply idw model for the data
# grab output of IDW for plotting
idw.output = as.data.frame(idw)  # output is defined as a data table
# set the names of the idw.output columns
# basic ggplot using geom_tile to display our interpolated grid within no map
ggplot() + 
  geom_tile(data = idw.output, aes(x = x, y = y, fill = var1.pred)) + 
  geom_point(data = mb, aes(x = x_WGS84, y = y_WGS84), shape = 21, color = "red") +
  scale_fill_distiller(palette = "PuOr", direction = 1) +
  theme_bw() 
```

### Add IDW results to `ggmap()`
To add the IDW results to the ggmap, the `geom_tile()` function is used to create a raster display of the interpolated values, the result of the `ggmap()` call is about the same.
```{r IDW_facet, comment='', cache=TRUE}
# use ggmap as earlier to display interpolation in geographic space
# Can take a LONG time to render (~1 minute)
ggmap(base_map, extent = "normal") +
  geom_tile(data = idw.output, aes(x = x, y = y, fill = var1.pred), alpha = 0.75) + 
  geom_point(data = mb, aes(x = x_WGS84, y = y_WGS84), shape = 19, color = "red") +
  scale_fill_distiller(palette = "PuOr", direction = 1) +
  theme_bw() 

```

## Interactive Mapping
Static maps are fantastic, but sometimes "slippy" interactive maps are great for presenting and exploring data.  The following code uses the `mapview` package to create the interactive maps.  This is a rapidly developing area for R mapping and other solutions currently exist, including [`leaflet`](https://rstudio.github.io/leaflet/).  

The process of making an interactive `mapview` map is pretty darn easy as we will see.  The first example is created simply by calling the `mapview()` function on the `germany` spatial object (imported from a shapefile)
```{r interactive_simple, comment=''}
mapview(germany)
```

#### Multi-layer map
Adding layers is a sample as creating a call to the `mapview()` function for each layer, and then adding them together.  Styling for each layer is done within the call to `mapview()`.
```{r interactive_layers, comment=''}
maplayer1 <- mapview(germany, 
                     color = "gray10", 
                     alpha.regions = 0)
maplayer2 <- mapview(mb_spat_WGS84)
maplayer1 + maplayer2
```

### Adding in the IDW raster
The first step to incorporate the IDW results into the interactive map is to turn it into a spatial object.  This is done with the function `rasterFromXYZ()` from the `raster` package.  Plot it using `ssplot()` to make sure it looks reasonable.
```{r IDW_raster, comment=''}
## use raster packaage to turn IDW into raster object
idw_raster <- rasterFromXYZ(idw.output[,1:3], crs = WGS84)
# plot to see what happens
spplot(idw_raster)
```

### Clip IDW raster
For this analysis, we only want to show the interpolation results within the area of which we have data points.  To do so, we will simply clip the raster.
To do so:
  1. Create clipping polygon from UTM projected feature location data
    + Find the envelope of the data points with `gConvexHull()`
    + Buffer that envelope to make our study region with `gBuffer()`
    + reproject to WGS84 for mapping
  2. Clip IDW raster
    + use `mask()` function in `raster` package to clip raster with buffer
  3. Create contours
    + use `rasterToContour()` function to create contours of IDW results
  
```{r buffer_clip, comment=''}
## To clip raster to data region, creat convex hull, buffer, and mask (clip) raster
## perform convex hull and buffer operations on projected UTM32n data
mb_hull <- gConvexHull(mb_spat) # convex hull
mb_buff <- gBuffer(mb_hull, width = 25000) # arbitrary 25km buffer
# tranforms back to WGS84
mb_buff_WGS84 <- spTransform(mb_buff, WGS84)
# plot to see results
plot(mb_buff_WGS84)
points(mb_spat_WGS84)
# mask IDW raster to confine to region that we have sites for
idw_raster_crop <- mask(idw_raster, mb_buff_WGS84)
# create contours of the IDW
idw_contour <- rasterToContour(idw_raster_crop, nlevels = 15)
# plot to see results
spplot(idw_raster_crop)
```

## Final map
Finally, all of the layers (Germany border, vessel locations, IDW raster, and contours) add added together as `mapview` objects.
```{r mapview, comment=''}
### Use mapview package to create layered interactive map with
### basemap, data points, interpolated raster, and contours
## made in three seperate layers
m1 <- mapview(germany, color = "black", alpha.regions = 0)
m2 <- mapview(mb_spat_WGS84, legend = TRUE, zcol = "new_type")
m3 <- mapview(idw_raster_crop, alpha.regions = 0.50, na.color = "transparent")
m4 <- mapview(idw_contour, lwd = 1.5, color = "gray40")
# combine layers to plot
m1 + m2 + m3 + m4
```





### Data Details

#### Description
A sites by types table of abundance data on vessel types in archaeological features of the
Younger Neolithic Michelsberg Culture from Belgium, France and Germany by Birgit Höhn (2002).

#### Details
Höhn (2002) recorded pottery vessel shapes from 108 archaeological features (pits, ditches etc.)
from 69 sites of the Central European Younger Neolithic Michelsberg Culture (MBK; 4350-3500 BC)
following Lüning’s (1967) typology. Her correspondence analysis of the abundance data (columns 5 to 39)
exhibits a pronounced Guttman effect or arch, suggesting the data set is structured by a time gradient.
Recently Mischka et al. (2015) projected an 109th Michelsberg assemblage, Flintbek LA48,
a pit with Michelsberg pottery from a North German site of the Funnel Beaker Cul- ture (TRB),
as a supplementary row into the existing chronology thereby connecting the relative chronologies
of TRB and MBK. The data frame contains as attributes the references for the data, a typological
key and the map projection. Note that ambiguous fragments of conical bowls (ks1 and ks2) are assigned
as 0.5 to each of the two types resulting also in positive entries suitable to analysis by CA.

#### Source
Höhn, B. 2002. Die Michelsberger Kultur in der Wetterau.
Universitätsforschungen zur prähis- torischen Archäologie 87. Bonn: Habelt.
Mischka, D., Roth, G. and K. Struckmeyer 2015. Michelsberg and Oxie in contact next to the Baltic Sea.
In: Neolithic Diversities. Perspectives from a conference in Lund, Sweden.
Acta Archaeologica Lundensia Ser. 8, No. 65, edited by Kr. Brink et al., pp 241–250.
Lüning, J. 1967. Die Michelsberger Kultur: Ihre Funde in zeitlicher und räumlicher Gliederung.
Berichte der Römisch-Germanischen Kommission 48, 1-350.