---
title: "SAA Seminar - R Basic Concept and Syntax"
author: "MDH"
date: "September 20, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R - Basic Concepts and Syntax
## Intro
Basic data types and working with data structures
R has three main data types: chracter, numeric, and logical
These data types are used as single objects, or within data structures
incuding: vector, matrix, list, and data.frame

## Characters

```{r characters, comment = ''}
# simple character
# "a" # RESULTS IN ERROR
# but not unquoted, because R expects that to be an object defined as somthing
# print(a) # returns: Error: object 'a' not found
# use the assignment operator '<-' to assign a value to the object 'a'
a <- "x"
print(a)
class(a) # tells us the data type of the object a: [1] "character"
# now the object 'a' is assigned the chracter value "x"

## That object can be used in a printed string with the paste0() function
print(paste0("object a contains the value: ", a))
# we can change the value of 'a' by reassigning it with another value
a <- "y" # R does not warn you that 'a' already is assigned a value!
print(paste0("object a contains the value: ", a))
```

## Numeric

```{r Load_data, comment = ""}
## object 'a' can also be a numeric
a <- 2
print(a)
# assign another number to object 'b'
b <- 5
# now we can add the two object since they are just placeholders for numbers
a + b
# this could also be accomplished simply by typeing:
2 + 5
# but using objects allows for the declaration of values, types, and all sorts of stuff

```

## Vectors

### Character vector
```{r chr_vec, comment = ''}
# character vector
v <- c("1", "x", "ch")
print(v)
v1 <- c("The", "brown", "dog")
print(v1)
v2 <- c("runs", "fast!")
print(v2)
# concatenate character vectors
v3 <- c(v1, v2)
# compose into sentence
new_sentence <- (paste0(v3, collapse = ' '))
print(new_sentence)
```

### Numeric vector

```{r num_vec, comment = ''}
# numeric vector 
bvec <- c(1,4,78)
print(bvec)
# logical vector
c <- c(TRUE, FALSE, TRUE)
print(c)
as.numeric(c)
```


### Numeric vector from sequences

```{r num_vec2, comment = ''}
# make a numeric vector from a sequence
bvec1 <- 1:10
# find the length of a vector
length(bvec1)
# or - to get the same thing
bvec2 <- seq(from = 1, to = 10, by = 1)
# add vectors
bvec1 + bvec2
# vector of boot strpped samples from vector b1
boot_samp <- sample(bvec1, 40, replace = TRUE)
print(boot_samp)
# a vector of length 1000 filled with random standard normals
set.seed(717)
rand_norm <- rnorm(1000,0,1)
length(rand_norm)
```


### Factors

```{r factors, comment = ''}
# make a factor data type 
colors <- sample(c("orange", "green", "blue"), 15, replace = TRUE)
fact1 <- factor(colors)
print(fact1)
# prints as color names, but is stored internally as integer values assigned based on alphabetical order
print(fact1)
levels(fact1)
as.numeric(fact1)
as.character(fact1)

## Sometimes the realtive order of the levels matter
fact2 <- factor(fact1, levels = c("blue", "orange", "green"))
levels(fact2)
## Other times, factors need to be expliclty ordered
ord_fact <- ordered(colors, levels = c("blue", "orange", "green"))
# notice that the Levels attribute shows that blue is less than orange is less than green
print(ord_fact)
```


### Qualifying and Quantifying vectors

```{r vec_functions, comment = ''}
table(v)
table(c)
table(boot_samp)
sum(boot_samp) 
mean(rand_norm)
var(rand_norm)
sd(rand_norm)^2
quantile(rand_norm)
```


### Creating a matrix

```{r matrices, comment = ''}
m1 <- matrix(1:20, nrow = 5)
print(m1)
m2 <- matrix(1:20, ncol = 6)
print(m2)
m2 <- matrix(1:20, ncol = 2)
print(m2)
m3 <- matrix(letters[1:20], ncol = 2)
print(m3)
dim(m3) # dimensions of matrix
nrow(m3) # number of rows in matrix
ncol(m3) # number of columns in matrix
summary(m3) # summary of values in columns
# change columns names
summary(m2) # default column names
colnames(m2) <- c("Column 1", "Column 2")
print(m2) # new column names
t(m2) # transposed matrix
# mixing char and num vectors results in all character matrix
m4 <- matrix(c(1:10,letters[1:10]), ncol = 2)
print(m4)
```

### Lists

```{r lists, comment = ''}
# ex. a list of three elements: 1) a single numeric, 2) a chracter vector, 3) a numeric marix
l1 <- list(b, new_sentence, m2)
print(l1)
y_var <- rbinom(10,1,0.5)
x_vars <- matrix(c(rnorm(10,0,1),rnorm(10,4,0.5)),ncol = 2)
mod1 <- glm(y_var ~ x_vars, family = "binomial")
model_list <- list(y = y_var, x = x_vars, model = summary(mod1))
print(model_list)
```

### indexing an subsetting

```{r index_vectors, comment = ''}
## Each of the data structures can be indexed and subsetted to retieve elements, rows, vectors, etc...
# for vectors
char1 <- v1[2]
print(char1)
num1 <- bvec[3]
print(num1)
num2 <- boot_samp[3:15]
print(num2)
```

### Indexing matrices

```{r index_matrix, comment = ''}
## Matrices are indexed by row (n) the column (m) as [n,m]
# single value from row 1, column 1
m2[1,1]
# single value from row 5, column 2
m2[5,2]
# get entire row - index by row number, but leave column index blank
# returns as single row matrix
m2[5,]
# get entire column - index by column number, but leave row index blank
# returns a vector
m2[,2]
# or return a single column matrix
m2[,2,drop = FALSE]
# return a range of rows, same works for columns
m2[1:3,]
```

### Apending a matrix
```{r append_matrix, comment = ''}
## adding rows or columns
# create some random data to append
new_row <- c(99,109)
new_col <- sample(1:11,11)
# add row with rbind() function
new_matrix <- rbind(m2,new_row)
print(new_matrix)
# add column with cbind() function
new_matrix <- cbind(new_matrix, new_col)
print(new_matrix)
# set row and column names
colnames(new_matrix) <- c("col1", "col2", "col3")
rownames(new_matrix) <- NULL
print(new_matrix)
```

### data.frames
data.frames are similar to matrices, but can store values of either num, char, or logic for each column
data.frames are more general than matrices and are a very common data format for analysis

```{r df1, comment = ''}
df1 <- data.frame(new_matrix)
print(df1)
## df1 contains only numeric data, but we can add a column of characters using cbind()
# a matrix would warn you about this and convert all data to characters
df1 <- cbind(df1, col4 = letters[1:nrow(df1)])
print(df1)
## It can be indexed the same way as a matrix using [row,column]
# returns a vector of the second column
df1[,2]
# or single values
df1[5,2]
## data.frames can also be indexed by column names in two different ways
# 1) quoted string with brackets as above, e.g. [,"column_name"]
# note that it was automatically converted to a factor when cbind() added it to the dataframe
df1[,"col4"]
# or index by the column as its own object
## or multiple columns using the c() function and column names
df1[ ,c("col1", "col4")]
# or 2) with the '$' operator 
df1$col1
```

### heading

```{r temp2, comment = ''}

```

### heading

```{r temp3, comment = ''}

```




### Data details

