---
title: "Untitled"
author: "MDH"
date: "September 20, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro
stuff

## Data setup
```{r Libraries, results = 'hide', message = FALSE}
### Load libraries
## These allow for extending the capabilities of base R
## Better to load all libraries up front
library("archdata") # for the archaeology data sets
library("dplyr") # for data manipulation
library("stringr") # for working with chracter strings
library("tidyr") 
```


## see how data is structured
```{r Load_data, comment = ""}
### Load data
## the 'DartPoints' dataset is included in the 'archdata' library loaded above
## Call the dataset we want by using the data() function
data(DartPoints)
# assign the dataset to a variable called 'dp' to save some typing
dp <- DartPoints

### Explore data
## These steps can lead you down many paths, but the idea is to look into the data
## to better understand what it includes and how it is structured
# the str() function prints out the strcture of the object
str(dp)
# dplyr package has a similar function
glimpse(dp)
# summary() function gives a summary of the data that varies by data type
summary(dp)
# head() function prints out the 'head' or first few occurences/rows of the data
head(dp)
# adding an integer as the second argument to head will instruct the function on how 
# many occurences/rows to print.  THe default is n = 5
head(dp, 10)
# the tail() function works the same way, but for the end of the dataset
tail(dp,10)
```

### Visualizing data
eee
```{r viz_data, comment = ''}
## Histogram
hist(dp$Length)
# hist() has additional arguments that control the bin count
hist(dp$Length, breaks = 30)
# boxplots allow for the visualization of aggregated distributions
boxplot(dp$Length)
# boxplots can also be grouped to compare distributions
boxplot(dp$Length ~ dp$Name)
# or bivariate combinations of all numric variables
pairs(dp[,c(7:13)])
```

####

```{r next, comment = ''}

```

### Data details
Details
Measurements on five types of dart points from Fort Hood in central Texas (Darl, Ensor, Pedernales, Travis, and Wells). The points were recovered during 10 different pedestrian survey projects during
the 1980's and were classified and measured by H. Blaine Ensor using the system created by Futato
(1983) as described in Carlson, S., et al 1987, pp 51-70 and Appendices 4 and 7.

Source
Fort Hood Projectile Points. Electronic database compiling the results of multiple surface surveys
at Fort Hood in the possession of David L. Carlson, Department of Anthropology, Texas A&M University,
College Station, TX. The artifacts are curated at Fort Hood, TX by the Cultural Resources
Branch of the Directorate of Public Works.

References
Carlson, S. B., H. B. Ensor, D. L. Carlson, E. A. Miller, and D E. Young. 1987. Archaeological
Survey at Fort Hood, Texas Fiscal Year 1984. United States Army Fort Hood. Archaeological
Resource Management Series, Research Report Number 14.
Futato, E. M. 1983. Projectile Point Morphology: Steps Toward a Formal Account. in Proceedings
of the Thirty-fourth Southeastern Archaeological Conference, Lafayette, Louisiana, October 27-19,
1977. Southeastern Archaeological Conference. Bulletin 21: 38-81.
