---
title: "SAA Seminar - R working with data"
author: "MDH"
date: "September 22, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basics of working with data in R:
# Operators, Loops, and Functions
R excels at letting users interactively work with data in a fluid manner.  This is achieved through an interactive (line by line) interpreter, data structures, and expressive syntax.  Using the basics of of data types, vectors, an subsetting, we can use the functions and methods presented here to mold data into the form needed for further analysis and modeling.  This is by no means an exhaustive or even representative overview of data processing functions/methods, but it will offer some of the fundamental aspects.

This overview is focused on functions in the base installation of R.  Another overview will briefly introduce data processing with in the ['Tidyverse'](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/); a set of packages and functions designed explicitly for  processing, tidying, and visualizing data in R.

Introduced here are:

* Operators
* Control structures
* Basic function
* Constructing User Defined Functions
* Apply functions

## Operators
R contains a series of mathematical, logical, relational, and other operators that are the foundation of working with data.  An [operator](https://www.tutorialspoint.com/r/r_operators.htm) is a typically single or double character function that serves to perform comparative or arithmetic operations on data object.  The most basic examples are `+` for additions, `>` or `<` greater than or less than to test relative magnitude, `==` for testing equivalency, and `<-` for assigning something to an object. Below are examples of a few of these.


### Arithmetic operators
R has all the functionality to work as a massively expanded calculator.  A wide list of Arithmetic operators can be used on numeric type objects.
```{r basic_operators, comment=''}
2 + 3 # addition
3 - 5 # subtraction
3 * 3 # multiplication
3^2 # raise to power of
34 %% 15 # modulo
sqrt(532) # square root
log(1003) # natural log
exp(34) # exponentiate
sum(1:10) # sum
mean(1:10) # mean
median(1:10) # median
var(1:10) # variance
```

#### Arithmetic operators on vectors and data frames
These functions are vectorized, meaning that they will automatically apply the function to each element of a vector. As such, they will return a vector of the same length
```{r arithmetic_examples, comment=''}
# build a data.frame to work on
set.seed(717)
df2 <- data.frame(col1 = rnorm(10,0,1),
                  col2 = rnorm(10,4,0.5),
                  col3 = sample(c(TRUE, FALSE),10,replace=TRUE),
                  col4 = letters[1:10],
                  stringsAsFactors = FALSE)
print(df2)
df2$col5 <- df2$col1 + df2$col2 # addition
print(df2$col5)
df2$col5 <- df2$col1 - df2$col2 # subtraction
print(df2$col5)
df2$col5 <- df2$col1 / df2$col2 # division
print(df2$col5)
df2$col5 <- df2$col1 * df2$col2 # multiplication
print(df2$col5)
df2$col5 <- df2$col1 %% 0.2 # modulo
print(df2$col5)
```

### Relational and Logical operators
Relational and Logical operators return a TRUE or FALSE if the expression evaluated to TRUE or FALSE.  These can be used to subset a vector or data.frame and may not return a logical vector of the same length as the input.
```{r realtional_operators, comment=''}
df2$col4 == "b" # equals (returns logical)
df2[which(df2$col4 == "b"),] # using which() to return columns
df2[which(df2$col1 < -0.5), "col1"] # less than
df2[which(df2$col1 >= -0.5), "col1"] # greater than or equal to
isTRUE(df2[1,"col3"]) # is a value TRUE
df2$col3 != TRUE # does not equal (returnd logical T/F)
df2$col4 %in% c("b","h","j") # within a set, returns logical
df2[which(df2$col4 %in% c("b","h","j")), "col4"] # returning values using which
df2[which(df2$col1 < -0.5 | df2$col1 > 0.5), "col1"] # element wise or
df2[which(df2$col1 < -0.5 & df2$col3 == TRUE), ] # element wise and
```


## Control structures
Control structures are a main tool in building a program.  While operators allow you to manipulate data one function at a time (or across a vector), [control structures](https://ramnathv.github.io/pycon2014-r/learn/controls.html) allow you to add a logical flow to the operations.  The basic structures demonstrated here common across most any computer language, although syntax will vary.  These structures can be nested to add increasing level of control (and complexity)  Grasping the utility of control structures is really important in making the conceptual leap from R as a high-powered calculator to R as an environment to building an analysis.  Once understood, even more powerful structures exist in R.

### Format of basic control structures
```{r, control_structures, eval = FALSE}
# if statement
if (condition) { 
  # do something
}
# if else statment
if (condition) {
  # do somthing
} else {
  # or do something else
}
# for loop
for (variable in vector) {
  # do something for each variable
}
# while loop
while (condition) {
  # do something while condition is TRUE
}
# nested loops
if (condition) {
    for (variable in vector) {
    # do something for each variable
  }
} else {
    if (condition) { 
    # do something
    } else {
    # do something else
  }
}
```

### Examples of control structures
```{r control_examples, comment=''}
# if statment
some_data <- rnorm(10,0,1)
if (mean(some_data) > 0) {
  some_new_data <- mean(some_data)
  print(mean(some_data))
}
# if else statment
some_data <- rnorm(10,0,1)
if (mean(some_data) > 0) {
  some_new_data <- mean(some_data)
  print(mean(some_data))
} else {
  some_new_data <- NULL
  print("Mean of data is less than zero")
}
# for loop
some_data <- rnorm(10,0,1)
for (i in 1:length(some_data)) {
  some_new_data <- some_data[i]^2
  print(some_new_data)
}
# nested ifelse in for loop
some_data <- rnorm(10,0,1)
some_new_data <- NULL
for (i in 1:length(some_data)) {
  iter_data <- some_data[i]
  if (iter_data > 0) {
    some_new_data[i] <- iter_data^2
    print("Data positive")
  } else {
    some_new_data[i] <- abs(iter_data)^2
    print("Data < 0, applied asb()")
  }
}
print(some_new_data)
# while loop # be careful
i <- 0
while(i < 4){
  new_value <- rbinom(1,1,0.5)
  i <- i + new_value
  print(paste0("Value of 'i' is: ", i))
}
```

### Vectorized version of nested loop
```{r vectorized_loop, comment=''}
# vectorized version of above
some_new_data2 <- ifelse(some_data > 0, some_data^2, abs(some_data)^2)
print(some_new_data2)
identical(some_new_data, some_new_data2)
```


## R functions to work with data
Both operators and control structures open the door to [functional programming](http://adv-r.had.co.nz/Functional-programming.html), but there are thousands of pre-build [functions](http://adv-r.had.co.nz/Functions.html) available in R and the package ecosystem, and an infinite number of User Defined Functions that you can create for your own needs. The real work of "learning R" is really to learn what functions exist, how they work, how to write your own, and how do so with efficiently and defensively.  By creating functions, you unlock the power and potential of R.

### Base functions for manipulating data
It is well beyond the scope of this class to cover the thousands of base functions, but a very small sample is presented.  The live demo and example rmarkdown files will demonstrate many more.

```{r base_functions, comment=''}
df2$col5 <- df2$col1 - mean(df2$col1)
print(df2$col5)
df2$col5 <- scale(df2$col1, center = TRUE, scale = FALSE)
print(df2$col5)
df2$col5 <- sign(df2$col1)
print(df2$col5)
df2$col5 <- ifelse(sign(df2$col1) == 1, TRUE, FALSE)
print(df2$col5)
df_sample <- sample(df2$col1, 20, replace = TRUE)
print(df_sample)
sort(df_sample)
rev(sort(df_sample))
```


### User Defined Function
Often in R, there is not available function to do what you want to do.  This may be because you are creating a new statistical method, that your need is very specific, or that you are extending the limits of an existing function, for example. In this case, you may create you own function.  Creating a function is very easy in R.  The structure of a new function is:
```{r function_structure, eval=FALSE}
name_of_new_function <- function(arguments){
  # do something here
  return(results)
}
```

Functions can have any name that follows convention, or no name at all (an anonymous function), can take any object type as input arguments, and can return any object type.  In the example below:

* I assign the object called `my_function` 
* the function that takes three arguments `x`, `y`, and `constant`
* the `ifelse()` evaluates if each $x_i$ is greater than or equal to zero
* if `TRUE`, $x_i$ is squared, if `FALSE` $x_i$ is divided by 2
* then the object `new_value` is assigned the `mean` of `y`
* multiplied by `x` (after being altered by the `ifelse()`)
* and the constant argument is added
* the values assigned to `new_value` is returned back to where the function was called.  In this case

Note: if any of the arguments are not of the type `numeric`, this will result in an error.  A more defensive function would include a test of the arguments that returns a helpful error message if that are anything but `numeric`.  Further, this function will work with either `numeric` vectors or single `numeric` values. However, the assignment back to `df2$col5` will result in an error because the length of the returned value is not equal to the length of the column vector it is being assigned to.

```{r my_function, comment=''}
my_function <- function(x, y, constant){
  x <- ifelse(x >= 0, x^2, x/2)
  new_value <- mean(y) * x + constant
  return(new_value)
}

df2$col5 <- my_function(x = df2$col1, y = df2$col2, constant = 0.5)
print(df2$col5)
```


One important thing to learn about functions is whats referred to as [scoping](http://www.programiz.com/r-programming/environment-scope), or where R looks for the variables you are using inside your function. Over simplifying a bit, the function will look for information in the arguments passed to it and the variables assigned within it; as opposed to looking outside the function (out into your program).  That means if you define the object `myVar <- 5` outside of the function, but do not pass it into the function, the function will not know what you mean by `myVar`.  Bottom line, while learning, do not use global variables.  If you want to use an object in a function, create it there or pass it in as an argument. A quick example:

```{r}
# function environment - look internally for myVar
inside_scoping <- function(x){
  myVar <- 1 # myVar defined inside function
  new_value <- x + myVar
  return(new_value)
}
myVar <- 5 # myVar also defined outside function
myValue1 <- inside_scoping(1)
print(myValue1) # function uses internal myVar

## or ##

# script environment - look globally for myVar
outside_scoping <- function(x){
  # myVar no longer defined in function, but used in here
  new_value <- x + myVar
  return(new_value)
}
myVar <- 5 # myVar only defined outside (globally)
myValue2 <- outside_scoping(1)
print(myValue2) # function finds myVar in the global environment and uses it
```


## `apply` functions
Putting the ideas of control structures, vectorizing, and functions together, we end up with the family of [apply functions in R](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family#gs.r6jOdjQ).  To paraphrase a common sentiment in R users, "If your first thought is 'how can I loop this', your second thought should be 'how can I vecotrize this instead'".  This sentiment is born from the fact that loops (for, each, if, etc...) can be slow and cumbersome compared to vectorized forms that take advantage of faster code and are more presentable (if you are comfortable reading them).  The `apply` functions are a first step at this more efficient version of R programming.  They take some getting used to, but can save you major headaches down the road.

The `apply` functions perform more complex operations to slice and aggregate data in vectors, matrices, and lists than the basic functions used earlier, and do so in faster and in fewer lines of code than making loops to do the job.  Here are some examples (see then link in the previous paragraph for a more in-depth treatment):

```{r apply_functions, comment=''}
# make a sample data.frame
df3 <- data.frame(col1 = rnorm(100,0,1),
                  col2 = rnorm(100,4,0.5),
                  col3 = rbinom(100,1,0.5))
# apply() # returns vector by applying function over margins of a matrix
column_means <- apply(df3,2,mean) # compute the mean of each column
print(column_means)
row_means <- apply(df3,1,mean) # compute the mean of each row
head(row_means)
sqrd_matrix <- apply(df3,1:2, function(x) x^2) # square each value in a matrix
head(sqrd_matrix)
# colMeans(), rowMeans(), colSums(), rowSums() stand in for the above
```

### The `by()` function performs operations by groups
```{r by_functions, comment=''}
# by # operations by a group
group_col_means <- by(df3[,1:2], df3$col3, colMeans) # column means by group
print(group_col_means)
```

### `lapply()` apply function to each element of a list; returns list
```{r other_apply_functions, comment=''}
l2 <- list(part1 = rnorm(5,0,1), part2 = rnorm(12,3,1))
print(l2)
list_means <- lapply(l2, mean)
print(list_means)
list_sums <- lapply(l2, sum)
print(list_sums)
# sapply - similar to lapply, but returns vector of matrix
list_means2 <- sapply(l2, mean)
print(list_means2)
### Other apply methods should you need them
```






