---
title: "Building a ggplot with Zubrow Data"
author: "MDH"
date: "September 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Many Iterations of a ggplot
This markdown is compiled to show the graphical output of conceptualizing and building a ggplot one step at a time.  Many of the steps here are very small iterations, but that is pretty much how it is done; deciding on one small change at a time.

This example is drawn from data in from Zubrow’s 1974 *Population Contact Climate New Mexican Pueblos*.  I enjoy this data set, so I have used it a few times.  So this overview [blog pots](https://matthewdharris.com/2016/03/19/visualizing-zubrow-ggplot2-cowplot/
) and [this post](https://matthewdharris.com/2016/08/12/ggplot2-step-by-step/
) that goes over the script here in detail.

The intention of this plot is to support an argument made by Zubrow that there was a 18th to 20th century population shift in Pueblo society based on cultural contact and climate change.  Of course it will take more that a single graphic to make this point, but a single graphic would go a long way to help a reader see it.  To that end, I first decided on a graphic that would be focused on the percent change from the date of the first data point at 1760.  I wanted to use percent change to scale the y axis a bit more evenly since some of Pueblos had total populations that dwarfed the smaller pueblos and decreased the visual impact of the trend. Further, period over period change was pretty variable and did not tell the story. Secondly, since there is a geographic component to this population shift, I need to arrange the Pueblo data groups geographically East to West.  Ideally this would be done a s single row, but there are too many sites for that to work.

#### Note about ggplot2 version
this relies on the development version of ggplot2 for the subtitle and caption text.  To install this version, first install the `devtools` package `install.packages("devtools")`, then install ggplot from github `devtools::install_github("hadley/ggplot2")`.  If you do not do this and have the CRAN version, everything should work except the subtitle and caption, and you will get a warning.


```{r packages, comment = ''}
library("ggplot2") # dev version
library("ggalt") # for ggsave()
library("extrafont")
```

```{r ggplot, comment = ''}
# read in data from CSV - you will need to set this path
wd <- "/Users/mattharris/Documents/R_Local/SAA_R_Demo_2016/SAA_R_Demo_2016"
dat <- read.csv(paste0(wd,"/data/Zubrow.csv"))

# need to grab these from the data.frame because that are used in some plots
pop <- dat$Population
year <- unique(dat$Year)

# base R line plot of population over year
plot(dat$Population ~ dat$Year, type = "l")

# ggplot line plot of population over year
ggplot(dat, aes(x = as.factor(Year), y = Population)) +
  geom_line()

# ggplot bar plot of population over year
ggplot(dat, aes(x = as.factor(Year), y = Population)) +
  geom_bar(stat = "identity")

# ggplot line plot of population over year, grouped by Site
ggplot(dat, aes(x = as.factor(Year), y = Population, group = Site)) +
  geom_line()

# same with smoothed lines
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
            se = FALSE, color = "gray10", size = 0.5)

# smoothed lines as a horizontal line at the mean
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "gray10", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35")

# facet wrap on the East_Label, each site is its own plot
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "gray10", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35") +
  facet_wrap( ~ East_label, nrow = 2)

# added background data into each facet
# adapted first geom_smooth to call the data and set the facet variable to NULL
# this way it plots all the data everytime, ignoring the facet call
# color the data in the second geom_smooth a brigth red to focus on that site's line
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(data = transform(dat, East_label = NULL),
                     method = "lm", formula = y ~ splines::bs(x, 3),
                     se = FALSE, color = "gray90", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35") +
  facet_wrap( ~ East_label, nrow = 2) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
                     se = FALSE, color = "red", fill = "gray70")

# Start to work on design by adding theme_bw() to remove most of the defaults
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(data = transform(dat, East_label = NULL),
              method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "gray90", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35") +
  facet_wrap( ~ East_label, nrow = 2) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "red", fill = "gray70") +
  theme_bw()

# Modify the axis lables, title, subtitle, and caption
# this require the developers version of ggplot2 and ggalt (both from Github)
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(data = transform(dat, East_label = NULL),
              method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "gray90", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35") +
  facet_wrap( ~ East_label, nrow = 2) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "red", fill = "gray70") +
  theme_bw() +
  labs(title="Population Change in New Mexico Pueblos, 1760 to 1950",
       subtitle="Arranged from East to West",
       caption="Data: Zubrow(1974)",
       x = "Year")

# added theme() block where all the detail work is done
# these lines adjust the facets
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(data = transform(dat, East_label = NULL),
              method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "gray90", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35") +
  facet_wrap( ~ East_label, nrow = 2) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "red", fill = "gray70") +
  theme_bw() +
  labs(title="Population Change in New Mexico Pueblos, 1760 to 1950",
       subtitle="Arranged from East to West",
       caption="Data: Zubrow(1974)",
       x = "Year") +
  theme(
    strip.background = element_rect(colour = "white", fill = "white"),
    strip.text.x = element_text(colour = "black", size = 7, face = "bold", 
                                family = "Trebuchet MS"),
    panel.margin = unit(0, "lines"),
    panel.border = element_rect(colour = "gray90")
  )

# Formatting axis text
# X axis text is rotate 90 degrees
# the changing of fonts requires the extrafont package and a bit of doing
# easier on OSX than Windows, but follow tutorial here:
# http://blog.revolutionanalytics.com/2012/09/how-to-use-your-favorite-fonts-in-r-charts.html
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(data = transform(dat, East_label = NULL),
              method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "gray90", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35") +
  facet_wrap( ~ East_label, nrow = 2) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "red", fill = "gray70") +
  theme_bw() +
  labs(title="Population Change in New Mexico Pueblos, 1760 to 1950",
       subtitle="Arranged from East to West",
       caption="Data: Zubrow(1974)",
       x = "Year") +
  theme(
    strip.background = element_rect(colour = "white", fill = "white"),
    strip.text.x = element_text(colour = "black", size = 7, face = "bold", 
                                family = "Trebuchet MS"),
    panel.margin = unit(0, "lines"),
    panel.border = element_rect(colour = "gray90"),
    axis.text.x = element_text(angle = 90, size = 6, family = "Trebuchet MS"),
    axis.text.y = element_text(size = 6, family = "Trebuchet MS"),
    axis.title = element_text(size = 8, family = "Trebuchet MS")
  )

# Finally adjust the title and caption text
# again, this requires the dev version of ggplot2 and galt
ggplot(dat, aes(x = as.factor(Year), y = log(Population), group = East)) +
  geom_smooth(data = transform(dat, East_label = NULL),
              method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "gray90", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35") +
  facet_wrap( ~ East_label, nrow = 2) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "red", fill = "gray70") +
  theme_bw() +
  labs(title="Population Change in New Mexico Pueblos, 1760 to 1950",
       subtitle="Arranged from East to West",
       caption="Data: Zubrow(1974)",
       x = "Year") +
  theme(
    strip.background = element_rect(colour = "white", fill = "white"),
    strip.text.x = element_text(colour = "black", size = 7, face = "bold", 
                                family = "Trebuchet MS"),
    panel.margin = unit(0, "lines"),
    panel.border = element_rect(colour = "gray90"),
    axis.text.x = element_text(angle = 90, size = 6, family = "Trebuchet MS"),
    axis.text.y = element_text(size = 6, family = "Trebuchet MS"),
    axis.title = element_text(size = 8, family = "Trebuchet MS"),
    plot.caption = element_text(size = 8, hjust=0, margin=margin(t=5), 
                                family = "Trebuchet MS"),
    plot.title=element_text(family="TrebuchetMS-Bold"),
    plot.subtitle=element_text(family="TrebuchetMS-Italic")
  )

## WITH HADLEY"S SUGGESTION OF DATES ON X AXIS
ggplot(dat, aes(x = Year, y = log(Population), group = East)) +
  geom_smooth(data = transform(dat, East_label = NULL),
              method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "gray90", size = 0.5) +
  geom_hline(yintercept = mean(log(pop)), linetype = 5, color = "gray35") +
  facet_wrap( ~ East_label, nrow = 2) +
  geom_smooth(method = "lm", formula = y ~ splines::bs(x, 3),
              se = FALSE, color = "red", fill = "gray70") +
  theme_bw() +
  scale_x_continuous(breaks = year, labels = year) +
  labs(title="Population Change in New Mexico Pueblos, 1760 to 1950",
       subtitle="Arranged from East to West",
       caption="Data: Zubrow(1974)",
       x = "Year") +
  theme(
    strip.background = element_rect(colour = "white", fill = "white"),
    strip.text.x = element_text(colour = "black", size = 7, face = "bold", 
                                family = "Trebuchet MS"),
    panel.margin = unit(0, "lines"),
    panel.border = element_rect(colour = "gray90"),
    axis.text.x = element_text(angle = 90, size = 6, family = "Trebuchet MS"),
    axis.text.y = element_text(size = 6, family = "Trebuchet MS"),
    axis.title = element_text(size = 8, family = "Trebuchet MS"),
    plot.caption = element_text(size = 8, hjust=0, margin=margin(t=5), 
                                family = "Trebuchet MS"),
    plot.title=element_text(family="TrebuchetMS-Bold"),
    plot.subtitle=element_text(family="TrebuchetMS-Italic")
  )
```